---
title: "FoodPriceIndex.Rmd"
author: "Matt Neibaur"
date: "7/27/2018"
output: html_document
---



```{r setup, include=FALSE,warning=FALSE, echo=FALSE, fig.height=6, fig.width=8}
knitr::opts_chunk$set(echo = TRUE)


#food price index, 5/2018, by Matt Neibaur, analysis of food prices
#Data from the United Nations: 
#The FAO Food Price Index is a measure of the monthly change in 
#international prices of a basket of food commodities.
#Relevance: During Irish famine, food prices spiked, food available but unafordable
#Wadhams, Peter. A Farewell to Ice: A Report from the Arctic (p. 141). 
#Oxford University Press. Kindle Edition. 
#Arab Spring could have been sparked by rising food prices
#web site: http://www.fao.org/worldfoodsituation/foodpricesindex/en/

#libraries used:
library(dplyr)
library(lubridate)
library(ggplot2)
#makes colors more separate
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

#functions
printGraph <- function(gName){
  ggsave(paste(output_folder,'/',gName, ".png",sep=""), width = 24, height = 16, units = "cm")
}

#location of files and folders:
code_folder <- "/Users/matt/R Code" #location of R code
input_folder <- "/Users/matt/R csvData"  #location of csv data
output_folder <- "/Users/matt/R Output"  #where output goes

###############
code_file = "FoodPrice1.R"
input_data = "Food_Price.csv"
###############

lookup = paste(input_folder,'/',input_data, sep="") #folder + file for lookup


#detach(FOA) #if re-reading file - make sure it is detached first
FOA <- read.csv(lookup, header=TRUE, quote="\"") 
#attach(FOA)

######################################


names(FOA)
# "Fdate"   "Food"    "Meat"    "Dairy"   "Cereals" "Oils"    "Sugar" 



```

## The Arab Spring 

* uprising may have been triggered by increased food prices
* first protests that occurred in Tunisia on 18 December 2010
* Reference: [Arab Spring](https://en.wikipedia.org/wiki/Arab_Spring)
* Preceded by riots in Africa in 2008: [Wikipedia](https://en.wikipedia.org/wiki/2007%E2%80%9308_world_food_price_crisis) 


According to [this independent news - *Food riots shut down main towns*:](http://www.irinnews.org/report/76905/burkina-faso-food-riots-shut-down-main-towns
)

>Riots over high food prices have already erupted in Burkina Faso’s Sahelian neighbours Mauritania and Senegal this year which are unusual in the region for being highly dependent on imported wheat and rice, products which have become more expensive worldwide this year. 

>In West Africa that situation is being compounded by unusually high food prices because of a disrupted growing season in some parts of the region in 2007 and reports that traders are hoarding stocks at markets in northern Nigeria, Ghana and elsewhere. 



```{r, include=TRUE,warning=FALSE, echo=FALSE}

df <- FOA %>%
  mutate(GGdate = mdy(Fdate)) %>%
  select(Food,GGdate) 

#Food price shock of 2007 & 2008, Burkina Faso riot
FirstRise <- df %>%
  filter(GGdate == "2008-02-20")


#first protests that occurred in Tunisia on 18 December 2010
#date for computer to match price entry, returns AS$GGdate, AS$Food
AS <- df %>%
  filter(GGdate == "2010-12-20")




#plot of dates vs food index
g <- ggplot(data=df)
g + geom_point(aes(x = GGdate, y = Food), alpha = 0.7, size = .5, col="black") +
    #geom_text(x=AS$GGdate, y=AS$Food,label = "x",size = 7,                               col="red", alpha = 0.7) +
    geom_point(x=AS$GGdate, y=AS$Food,size = 2, col="red",alpha = 0.7) +
    geom_point(x=FirstRise$GGdate, y=FirstRise$Food,size = 2, col="red",alpha = 0.7) +
  
    geom_smooth(aes(x = GGdate, y = Food), method = lm, alpha = 0.7, size = 0.5, col = "black") +
    labs(title = "Food Price Index", 
       subtitle = "Burkina Faso riot: Feb 2008, Arab Spring start: Dec 18, 2010 (red dot)",
       caption = "Source: United Nations FAO",
       x="Date", y = "Food Price Index")
  printGraph("FAO.food")
  

df <- FOA %>%
  mutate(GGdate = mdy(Fdate)) %>%
  select(GGdate,Meat) 

#Food price shock of 2007 & 2008, Burkina Faso riot
FirstRise <- df %>%
  filter(GGdate == "2008-02-20")


AS <- df %>%
  filter(GGdate == "2010-12-20")




#plot of dates vs food index
g <- ggplot(data=df)
g + geom_point(aes(x = GGdate, y = Meat), alpha = 0.7, size = .5, col="black") +
    geom_point(x=AS$GGdate, y=AS$Meat,size = 2, col="red",alpha = 0.7) +
    geom_point(x=FirstRise$GGdate, y=FirstRise$Meat,size = 2, col="red",alpha = 0.7) +
  geom_smooth(aes(x = GGdate, y = Meat), method = lm, alpha = 0.7, size = 0.5, col = "black") +
    labs(title = "Food Price Index (Meat)", 
       subtitle = "Burkina Faso riot: Feb 2008, Arab Spring start: Dec 18, 2010 (red dot)",
       caption = "Source: United Nations FAO",
       x="Date", y = "Food Price Index (Meat)")
  printGraph("FAO.Meat")
  
  


df <- FOA %>%
  mutate(GGdate = mdy(Fdate)) %>%
  select(GGdate,Dairy) 

#Food price shock of 2007 & 2008, Burkina Faso riot
FirstRise <- df %>%
  filter(GGdate == "2008-02-20")


AS <- df %>%
  filter(GGdate == "2010-12-20")




#plot of dates vs food index
g <- ggplot(data=df)
g + geom_point(aes(x = GGdate, y = Dairy), alpha = 0.7, size = .5, col="black") +
    geom_point(x=AS$GGdate, y=AS$Dairy,size = 2, col="red",alpha = 0.7) +
    geom_point(x=FirstRise$GGdate, y=FirstRise$Dairy,size = 2, col="red",alpha = 0.7) + 
  geom_smooth(aes(x = GGdate, y = Dairy), method = lm, alpha = 0.7, size = 0.5, col = "black") +
    labs(title = "Food Price Index (Dairy)", 
       subtitle = "Burkina Faso riot: Feb 2008, Arab Spring start: Dec 18, 2010 (red dot)",
       caption = "Source: United Nations FAO",
       x="Date", y = "Food Price Index (Dairy)")
  printGraph("FAO.Dairy")
  
  



df <- FOA %>%
  mutate(GGdate = mdy(Fdate)) %>%
  select(GGdate,Cereals) 

#Food price shock of 2007 & 2008, Burkina Faso riot
FirstRise <- df %>%
  filter(GGdate == "2008-02-20")


AS <- df %>%
  filter(GGdate == "2010-12-20")




#plot of dates vs food index
g <- ggplot(data=df)
g + geom_point(aes(x = GGdate, y = Cereals), alpha = 0.7, size = .5, col="black") +
  geom_point(x=AS$GGdate, y=AS$Cereals,size = 2, col="red",alpha = 0.7) +
  geom_point(x=FirstRise$GGdate, y=FirstRise$Cereals,size = 2, col="red",alpha = 0.7) +
    geom_smooth(aes(x = GGdate, y = Cereals), method = lm, alpha = 0.7, size = 0.5, col = "black") +
    labs(title = "Food Price Index (Cereals)", 
       subtitle = "Burkina Faso riot: Feb 2008, Arab Spring start: Dec 18, 2010 (red dot)",
       caption = "Source: United Nations FAO",
       x="Date", y = "Food Price Index (Cereals)")
  printGraph("FAO.Cereals")
  



df <- FOA %>%
  mutate(GGdate = mdy(Fdate)) %>%
  select(GGdate,Oils) 

#Food price shock of 2007 & 2008, Burkina Faso riot
FirstRise <- df %>%
  filter(GGdate == "2008-02-20")

AS <- df %>%
  filter(GGdate == "2010-12-20")




#plot of dates vs food index
g <- ggplot(data=df)
g + geom_point(aes(x = GGdate, y = Oils), alpha = 0.7, size = .5, col="black") +
    geom_point(x=AS$GGdate, y=AS$Oils,size = 2, col="red",alpha = 0.7) +
    geom_point(x=FirstRise$GGdate, y=FirstRise$Oils,size = 2, col="red",alpha = 0.7) +
  geom_smooth(aes(x = GGdate, y = Oils), method = lm, alpha = 0.7, size = 0.5, col = "black") +
    labs(title = "Food Price Index (Oils)", 
       subtitle = "Burkina Faso riot: Feb 2008, Arab Spring start: Dec 18, 2010 (red dot)",
       caption = "Source: United Nations FAO",
       x="Date", y = "Food Price Index (Oils)")
  printGraph("FAO.Oils")
  
  


df <- FOA %>%
  mutate(GGdate = mdy(Fdate)) %>%
  select(GGdate,Sugar) 

#Food price shock of 2007 & 2008, Burkina Faso riot
FirstRise <- df %>%
  filter(GGdate == "2008-02-20")


AS <- df %>%
  filter(GGdate == "2010-12-20")




#plot of dates vs food index
g <- ggplot(data=df)
g + geom_point(aes(x = GGdate, y = Sugar), alpha = 0.7, size = .5, col="black") +
    geom_point(x=AS$GGdate, y=AS$Sugar,size = 2, col="red",alpha = 0.7) +
    geom_point(x=FirstRise$GGdate, y=FirstRise$Sugar,size = 2, col="red",alpha = 0.7) +
  geom_smooth(aes(x = GGdate, y = Sugar), method = lm, alpha = 0.7, size = 0.5, col = "black") +
    labs(title = "Food Price Index (Sugar)", 
       subtitle = "Burkina Faso riot: Feb 2008, Arab Spring start: Dec 18, 2010 (red dot)",
       caption = "Source: United Nations FAO",
       x="Date", y = "Food Price Index (Sugar)")
  printGraph("FAO.Sugar")
  
  
```



